---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setoptions}

library(knitr)
opts_chunk$set(echo = TRUE, results = "asis")

```

## Loading and preprocessing the data

```{r preprocess}

if(!file.exists("activity.csv")){
    unzip("activity.zip")
}

data <- read.csv("activity.csv")

```

## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day

```{r stepsperday}

groupdata<- aggregate(data$steps, list(data$date), sum, na.rm=TRUE)
kable(groupdata, 
      col.names = c("Date", "Total Steps"), align = 'l', 
      caption = "Total Steps Per Day")

```

2. Make a histogram of the total number of steps taken each day

```{r histogramsteps}

hist(groupdata$x, xlab="Total Steps", main="Histogram of Total Steps")

```

3.Calculate and report the mean and median of the total number of steps taken per day

```{r meanmediansteps}

groupdata <- aggregate(data$steps, list(data$date), mean, na.rm=TRUE)
kable(groupdata, 
      col.names = c("Date", "Average Steps"), align = 'l', 
      caption = "Avg. Steps Per Day")

groupdata <- aggregate(data$steps, list(data$date), median, na.rm=TRUE)
kable(groupdata, 
      col.names = c("Date", "Median Steps"), align = 'l', 
      caption = "Median Steps Per Day")

```

## What is the average daily activity pattern?

1. Make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days

```{r stepsperinterval}

groupdata <- aggregate(data$steps, list(data$interval), mean, na.rm=TRUE)
with(groupdata, plot(Group.1, x, type = "l", xlab = "Interval", 
                     ylab = "Average Steps", 
                     main = "Average Steps Per Interval"))

```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r maxstepsperinterval}

print(paste("Interval ", groupdata[which.max(groupdata$x),1], " is the maximum interval with average steps of ", groupdata[which.max(groupdata$x),2]))

```

## Imputing missing values

1. Calculate and report the total number of missing values in the dataset

```{r missingvalues}

missing <- is.na(data$steps)
print(paste("Number of missing values = ", sum(missing)))

```

2. Devise a strategy for filling in all of the missing values in the dataset.

Strategy used is to fill the missing values with the mean for that 5-minute interval

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r imputingvalues}

imputdata <- data
names(groupdata) <- c("interval", "steps")
imputdata <- merge(imputdata, groupdata, by = "interval")
imputdata[is.na(imputdata$steps.x), ]$steps.x <- 
    imputdata[is.na(imputdata$steps.x),]$steps.y

```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

Based on the graphs below, there is a significant difference from using the original data versus the data with imputted valeus, causing an increase with the total number of steps.

```{r meanmedianimputsteps}

groupdata<- aggregate(imputdata$steps.x, list(imputdata$date), sum, na.rm=TRUE)
hist(groupdata$x, xlab="Total Steps", main="Histogram of Total Steps")

groupdata <- aggregate(imputdata$steps.x, list(imputdata$date), mean, na.rm=TRUE)
kable(groupdata, 
      col.names = c("Date", "Average Steps"), align = 'l', 
      caption = "Avg. Imputed Steps Per Day")

groupdata <- aggregate(imputdata$steps.x, list(imputdata$date), median, na.rm=TRUE)
kable(groupdata, 
      col.names = c("Date", "Median Steps"), align = 'l', 
      caption = "Median Imputed Steps Per Day")

```

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r dayofweekpattern}

imputdata$factor <- ifelse(weekdays(as.Date(imputdata$date))%in%c("Saturday", "Sunday"), "Weekend", "Weekday")

```

2. Make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days

```{r dayofweekplot}

library(ggplot2)
groupdata <- aggregate(data$steps, list(data$interval), mean, na.rm=TRUE)
g <- ggplot(imputdata, aes(interval, steps.x)) + labs(x = "Interval", y = "Average Imputted Steps", title = "Average Imputted Steps Per Interval")
g  + stat_summary(fun.y = mean, geom="line", size=1) + facet_grid(factor~.) 
    

```